Influence of Time Integration Methodology on Double Detonation Type Ia Supernova Nucleosynthesis

### 2023-06-13 subch_tols/paper.tex
start of subch tol paper

@@

\begin{document}
%======================================================================
% Title
%======================================================================
\title{Comparing Early Evolution of Flames in X-ray Bursts in Two and Three Dimensions}

\shorttitle{Early XRB Flame Evolution}
%% \shortauthors{Harpole et al.}

\author[0000-0001-8401-030X]{Michael Zingale}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

\author[0000-0001-6191-4285]{Kiran Eiden}
\affiliation{Department of Astronomy,
University of California, Berkeley,
Berkeley, CA 94720-3411, USA}

\author[0000-0003-0439-4556]{Max Katz}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

@@

X-ray bursts (XRBs) result from thermonuclear burning of an accreted
H/He or He layer on a neutron star \citep{galloway:2017}.
Observations of brightness oscillations in the rise of the lightcurve
suggest that the burning begins localized and then spreads across the
neutron star \citep{bhattacharyya:2007}. 
This spreading is inherently
a multi-dimensional phenomenon, requiring hydrodynamic simulations that
resolve the reactive zone, realistic reaction networks, and
domain sizes that capture the scales over which rotation is important in order to
understand the flame propagation and nucleosynthesis.  Detailed
nuclear physics is especially important for understanding the 
rp-process in mixed H/He bursts, as discussed in \citet{rpprocess}.

Here we show a first attempt at modeling the early evolution of a spreading
hotspot in three dimensions.  This builds on our earlier two-dimensional
work \citep{eiden:2020,harpole:2021} that developed our
simulation framework and explored how the acceleration of the burning
front depends on the initial model structure.  As with those
simulations, we use the freely-available \castro\ simulation code
\citep{castro,castro_joss}, with all of the code needed to run these
simulations in our public GitHub repositories.

Our simulations complement the two- and three-dimensional models of flames of
\citet{cavecchi:2013,art-2015-cavecchi-etal,art-2016-cavecchi-etal,Cavecchi2019}
by focusing on resolving the reaction zone of the flame and modeling
the vertical structure hydrodynamically instead of hydrostatically.
We also focus on modeling flames starting from a small hot spot.  This
however means that we are confined to smaller domains, with similar
computational resources, so one of the goals of this study is to
understand how far we can push resolved XRB flame simulations.  The
work by \citet{goodwin:2021} also explores multidimensional evolution,
looking at the thermal transport and how \edit1{it} affects the location where
the hotspot ignites.  Together, all three approaches help build a
comprehensive multidimensional picture of X-ray bursts.

The goal of this paper is to understand the challenges of resolved
three-dimensional hydrodynamical simulation and explore what is
possible with today's resources.  This will inform our follow-on
simulations.

@@

In \citet{eiden:2020}, we developed the simulation methodology in
\castro\ for modeling laterally spreading flames through the accreted
layer on the surface of a neutron star.  For these models,
\castro\ evolves the compressible Euler equations with a Coriolis
force arising from rotation, constant gravity (the plane-parallel
approximation), reactive sources, and thermal diffusion (integrated
explicitly)\edit1{:
\begin{eqnarray}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &=& 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho {\bf g} - 2 \rho {\bf \Omega} \times \Ub \\
%% no centrifugal- \rho {\bf \Omega} \times ({\bf \Omega} \times {\bf r}) \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot (\rho \Ub E + p \Ub) &=& \rho \Ub \cdot {\bf g} + \nabla \cdot k_\mathrm{th} \nabla T + \rho \dot{\epsilon}
%% no centrifugal - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho | \Omegab |^2 (\Ub \cdot r)
\end{eqnarray}
Here, $\rho$ is the mass density, $X_k$ are the mass fractions of the
nuclear species with associated creation rates $\omegadot_k$, $\Ub$
the velocity, $E$ is the specific total energy, $p$ is the pressure,
${\bf g}$ is the gravitational
acceleration, $\Omegab$ is the rotation vector, $k_\mathrm{th}$ is the thermal conductivity, and $T$ is the temperature.
}
 We use a general stellar equation of state with
radiation, ions as a perfect gas, and degenerate/relativistic
electrons described in \citet{timmes_swesty:2000} \edit1{which closes
  the system:}
\begin{eqnarray}
E &=& e + \frac{1}{2} |\Ub|^2 \\
p &=& p(\rho, e, X_k) \\
T &=& T(\rho, e, X_k)
\end{eqnarray}
Building on these
first simulations, in \citet{harpole:2021}, we explored several
different rotation rates and initial model thermal structures to
understand how the flame speed depends on the thermal structure of the
envelope.  Both of these papers used a two-dimensional axisymmetric
geometry.  The three-dimensional simulations we present here build on
these works.  We use a neutron star crust temperature of $2\times
10^8~\mathrm{K}$ and rotation rate of 1000~Hz, as followed in
\citet{harpole:2021}.  This model was chosen here since it gives a
clean well-defined flame.  The model is constructed following the same
procedure described in \citet{eiden:2020,harpole:2021}.  \edit1{We
generate a hot and a cool model, and blend them laterally to setup
an initial perturbation that is in hydrostatic equilibrium.  The
structure of these models is shown
shown in Figure~\ref{fig:initial model}.}

\begin{figure}[t]
\centering
\plotone{initial_model_paper}
\caption{\label{fig:initial model} Initial model structure for both
the hot (dashed) and cool (solid) models, showing the density
and temperature structure as well as the initial composition.}
\end{figure}

Moving to 3D is very expensive, so to further reduce the computational
expense, we use an anisotropic resolution, with 32 cm resolution
laterally and 16 cm resolution vertically.  This is finer vertically
than done in \citet{harpole:2021}.  Because the hotspot is placed in
the center of the domain, the size of the domain also becomes a
constraint, and as a result, the evolution time we can reach will
become limited by the time it takes the burning to approach the edge
of the domain.  Consequently, we focus here on the early evolution.

\castro\ uses the the \amrex\ adaptive mesh refinement library
\citep{amrex_joss} to manage the discretization and domain
decomposition.  Our 3D simulation used a base grid of $768^2 \times
192$ zones with 2 levels of refinement, the first jumping by a factor
of 4 and the next by a factor of 2.  We subcycle in time, so the
coarser grids take a large timestep than the fine grids.  We used
static mesh refinement, to achieve better load balancing, fully
refining the atmosphere below a height of $3600~\mathrm{cm}$.  The
domain has a size $(1.96608\times 10^5~\mathrm{cm})^2 \times
2.4576\times 10^4~\mathrm{cm}$.  The simulation methodology and
parameters are mostly identical to those used in \citet{harpole:2021}.
We use an unsplit piecewise parabolic method
\citep{ppmunsplit,millercolella:2002} for the advection and operator
(Strang) splitting for the reactions, evolving an internal energy
evolution equation during the burn, as described in
\citet{strang_rnaas}.  These simulations used the 7-isotope He-burning
network described in \citet{iso7}.  This was one of the networks used
in \citet{eiden:2020}.  \edit1{Thermal diffusion is treated explicitly, using the conductivities from \citet{Timmes00}.
The overall integration strategy is second-order accurate in space and time.}  The main difference with the previous
simulations is that we switched from a hydrostatic to reflecting
boundary at the bottom of the domain and used a simple well-balanced
scheme (similar to \citealt{kappeli:2016} but adapted to deal with
characteristic tracing in PPM as described in \citealt{ppm-hse}).
Tests demonstrated that this boundary does a better job than our
previous approach at supporting the atmosphere as the flame propagates
for long times.

The simulations were run on the OLCF Summit supercomputer, on 342 to
1366 nodes, with 6 NVIDIA V100 GPUs per node.  The entire computation
was offloaded to GPUs using the \amrex\ C++ abstraction layer for parallel \texttt{for} loops, as described in
\citet{castro_gpu}.  GPU offloading was critical to being able to
perform these simulations, since we run more than an order of
magnitude faster using all the GPUs on a node as compared to all of
the CPU cores on the node.  Overall, about 250,000 node-hours were used for
the calculation.  The full state of the calculation was output only
every 0.005 s, as each output file is 2.4 TB in size (for single
precision data).  We output a few fields more frequently for
visualization, and global diagnostics were output every timestep.  We
ran a 2D simulation with the same resolution and refinement strategy
to compare with here.

\begin{figure}[t]
\centering
\plotone{full}
\caption{\label{fig:vr_abar} Volume rendering of the three-dimensional XRB
simulation showing the mean molecular weight, $\bar{A}$.  \edit1{Two views
are shown: the left column is viewed from above while the right column
shows it from a shallow angle above the surface.  The panels show the state at 10 ms, 20 ms, and 40 ms (from top to bottom).
The colorbar shows the value of $\bar{A}$ together with the opacity assigned for
that value---this allows us to bring out some structure in the rendering.  A
small triad in the upper right of the top panels shows the orientation,
with red = $x$, green = $y$, and blue = $z$, our vertical direction.}}
\end{figure}

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\frac{1}{\bar{A}} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken\edit1{, likely due to roundoff error,}  and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.
\begin{figure*}[t]
\centering
\plotone{time_series}
\caption{\label{fig:2d_abar} Time-sequence of the two-dimensional axisymmetric
simulation using the same setup as the three-dimensional simulation shown
in Figure~\ref{fig:vr_abar}.  The domain is zoomed in, showing only the left
two-thirds of the surface.}
\end{figure*}

\begin{figure*}[t]
\centering
%\epsscale{0.75}
\plotone{time_series_3D}
%\epsscale{1.0}
\caption{\label{fig:3dslice} Vertical slices through the three-dimensional
simulation \edit1{at the same times as in Figure~\ref{fig:2d_abar}}, showing a region centered on the flame viewed from the side.  The
flame structure is clearly seen. }
\end{figure*}

Figure~\ref{fig:2d_abar} shows the corresponding two-dimensional
axisymmetric simulation.  In the axisymmetric geometry, each zone is
essentially a torus, rotated about the vertical axis.  The initial
model and simulation parameters are identical to the three-dimensional
case.  We see that the overall structure and evolution looks identical
to that reported in \cite{harpole:2021}.  We also see that there is a
well-defined flame in the simulation plane that spreads outward with
time.  A vertical slice through the three-dimensional simulation (in
the $x$-$z$ plane) is shown in Figure~\ref{fig:3dslice} at the final
time, 40~ms.  Aside from the lack of reflection symmetry from the
absence of axisymmetric geometry, it looks very similar to the two-dimensional case.

%% \begin{figure}[t]
%% \centering
%% \plotone{flame_wave_1000Hz_25cm_smallplt88081_Hnuc_annotated_top} \\
%% \plotone{flame_wave_1000Hz_25cm_smallplt174562_Hnuc_annotated_top.png}
%% \caption{\label{fig:vr_hnuc} stuff}
%% \end{figure}

Because the 3D flame does not stay perfectly circular, it is difficult
to define the flame speed using the same method employed in
\citet{eiden:2020}.  Instead, we will look at the mass of the ash
material to assess how quickly the burning is taking place.
Figure~\ref{fig:mass_plot} shows the mass of the different species as
a function of time, for the 2D and 3D calculation.  The axisymmetric
2D calculation has slightly less volume than the 3D domain, since
rotating about the symmetry axis produces a cylindrical domain
inscribed in the 3D domain.  To compensate for this, we scale the
masses by the total mass on the grid.  This plot shows that the
masses of the heavy species, especially \isot{C}{12}, grow quickly
with time.  The differences between the 2D and 3D simulations appear
small---the burning is slightly faster in 2D, but the trends track
very well.  This slight difference is likely because the assumption of
axisymmetry there does not allow for the complex structure we see in
the evolution of composition in the 3D flame.  This strong agreement
suggests that using 2D axisymmetric calculations is a good model for
the early flame evolution.  And since they are so much less
computationally expensive, this will allow us to explore much more
complex reaction networks and understand the nucleosynthesis in more
detail.

\begin{figure}[t]
\centering
\plotone{mass_plot}
\caption{\label{fig:mass_plot} Mass of species scaled to \edit1{total}
  mass as a function of time.  The 3D simulation is shown as the solid
  lines and the 2D simulation is shown as the dashed lines.}
\end{figure}

@@

We have extended our simulation framework to three-dimensions, and
explored the differences between a two-dimensional axisymmetric
simulation and a fully three-dimensional hydrodynamical simulation.
Overall, the early evolution of the flame spreading from a hotspot
behaves similarly in two- and three-dimensions.  As this
three-dimensional simulation pushed the limits of available
computational resources, this result suggests that we can continue to
use two-dimensional axisymmetric simulations to explore the initial
flame structure and spend our computational resources on larger networks
and bigger domains.

These results also suggest that we can do initial explorations of full
star bursts in two-dimensions, perhaps using axisymmetric spherical
coordinates (modeling an annular region in \edit1{neutron star radius and colatitude}).  In this
geometry, point ignition requires starting at the poles, but this
would allow us to see the effect of varying gravity with latitude (due
to centrifugal forces), as well as begin to consider lightcurve modeling.

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernova, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\citep{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accurate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in \castro~\citep{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.

We are currently exploring mixed H/He bursts as well as the
sensitivity of the flame properties to the size of the reaction
network used.  Both of these initial studies use our two-dimensional
axisymmetric geometry.

>>>>

\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddxs}[1]{{\frac{{\partial}}{\partial x}}#1}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

\newcommand{\Ic}{{\boldsymbol{\mathcal{I}}}}
\newcommand{\Ics}{{\mathcal{I}}}
\newcommand{\kmax}{{k_\mathrm{max}}}
\usepackage{bm}

\newcommand{\Uc}{{\,\bm{\mathcal{U}}}}
\newcommand{\Fb}{\mathbf{F}}
\newcommand{\Sc}{\mathbf{S}}

\newcommand{\xv}{{(x)}}
\newcommand{\yv}{{(y)}}
\newcommand{\zv}{{(z)}}

\newcommand{\ex}{{\bf e}_x}
\newcommand{\ey}{{\bf e}_y}
\newcommand{\ez}{{\bf e}_z}

\newcommand{\Ab}{{\bf A}}
\newcommand{\Sq}{{\bf S}_\qb}
\newcommand{\Sqhydro}{{\Sq^{\mathrm{hydro}}}}
\newcommand{\qb}{{\bf q}}

\newcommand{\Gb}{{\bf G}}
\newcommand{\Rb}{{\bf R}}
\newcommand{\Rq}{{\bf R}}
\newcommand{\Adv}[1]{{\left [\boldsymbol{\mathcal{A}} \left(#1\right)\right]}}
\newcommand{\Advt}[1]{{\left [\boldsymbol{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advss}[1]{{\left [{\mathcal{{A}}} \left(#1\right)\right]}}
\newcommand{\Advsst}[1]{{\left [{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advs}[1]{\boldsymbol{\mathcal{A}} \left(#1\right)}

@@

\begin{document}
%======================================================================
% Title
%======================================================================
\title{Influence of Time Integration Methodology on Double Detonation Type
Ia Supernova Nucleosynthesis}

\shorttitle{Integration Methods and SNe Ia Nucleosynthesis}

\author[0000-0001-8401-030X]{Michael Zingale}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800, USA}

\author[0000-0002-2839-107X]{Zhi Chen}
\affiliation{Department of Physics and Astronomy,
Stony Brook University,
Stony Brook, NY 11794-3800, USA}

\author[0000-0003-0439-4556]{Max Katz}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800, USA}

\author[0000-0001-5961-1680]{Alexander Smith Clark}
\affiliation{Department of Physics and Astronomy,
Stony Brook University,
Stony Brook, NY 11794-3800, USA}

\author[0000-0003-3603-6868]{Eric T. Johnson}
\affiliation{Department of Physics and Astronomy,
Stony Brook University,
Stony Brook, NY 11794-3800, USA}

@@

The double detonation model

Detonations can be difficult to model numerically.  Time step cuts \MarginPar{ref}.
For the double detonation scenario, the ignition of the second detonation
can be very sensitive to how the simulation is run \MarginPar{refs}.
Artifically limiting the rates \MarginPar{ref}

Integration of reaction networks takes various forms in astrophysical
simulation codes.

When using operator-splitting, the density remains constant, since there
is no reaction source in the mass continuity equation.  The mass
fractions, $X_k$, evolve according to:
\begin{equation}
\frac{dX_k}{dt} = \omegadot_k
\end{equation}
where $\omegadot_k$ is the creation rate of species $k$ due to nuclear
reaction.  In the simplest approximation of operator splitting, the
evolution of the mass fractions is solved alone, without integrating
the temperature or energy.  This is the form that was used in FLASH
\cite{flash}.  A more accurate operator splitting also includes the
energy evolution (perhaps in terms of temperature).  This will require
calling the equation of state each time we evaluate the righthand side
of the ODE system, increasing the cost.  However, as we showed in
\citet{strang_rnaas}, this is needed to get second-order convergence.

In \citet{castro_simple_sdc}, we introduced a method based on the
ideas of spectral deferred corrections for coupling reactions and
hydrodyamics (that we called ``simplified-SDC'').  Here, the overall
time-integration is done iteratively, with the hydrodynamics seeing an
explicit reactive source, and the reaction update evolving an ODE
system that includes a piecewise-constant-in-time advective source,
$\Adv{\Uc}^{n+1/2}$,
\begin{equation}
\ddt{\Uc} = \Rb(\Uc) + \Adv{\Uc}^{n+1/2}
\end{equation}

In addition to the time-integration strategy, integrating the reaction
ODE system requires specifying tolerances.  Unfortunately, the tolerances
used are not normally reported in papers, and in many cases, may not
be easily controlled by a code user at runtime.

\citet{castro_simple_sdc} looked at an extreme version of the 
double detonation scenario---a very large perturbation was applied
to drive a detonation directly into the underlying 
carbon-oxygen white dwarf.  We showed in that paper that the Strang
split integration method had difficulty with the integration unless
we used tighter tolerances

Summary of the SDC subch results form the original paper

@@

We use a 2d axisymmetric domain with a size of $5.12\times 10^9~\mathrm{cm}$ by $1.024\times 10^{10}~\mathrm{cm}$.  This is
much larger than the initial size of the star, giving it plenty of
room to expand.  The coarse grid is $640\times 1280$ zones and we use
2 levels of refinement, giving a maximum resolution of
$20~\mathrm{km}$.  The refinement strategy is picked to refine on the
star and any regions where the temperature is greater than
$10^8~\mathrm{K}$.  \castro\ uses subcycling, so finer grids are
advanced at a smaller timestep than the coarse grids.  Self-gravity is
done using a full Poisson solve using geometric multigrid, with
Dirichlet boundary conditions on the domain boundary computed via a
multipole expansion with a maximum order of 6.  For the low density
regions outside of the star, we use a sponge term on the momentum
equation to prevent the very low density material that is not in
hydrostatic equilibrium from raining down on the star.

If during the advance of a timestep an error is generated (negative
density, ODE integration fails, mass fractions don't sum to 1), the
step is thrown-out and retied with a smaller timestep.  This is done
on a level-by-level basis in the overall AMR subcycling hierarchy.

The reaction system is integrating using a version of the VODE ODE integrator
\citep{vode} ported to C++.  Our modifications to this integrator are described
in \citet{castro_simple_sdc}.

We will run a suite of simulations all using the same initial model.
The initial model is exactly the same as in \citep{castro_simple_sdc}.

The simulations are all run on the OLCF Frontier machine, using 4
nodes / 32 AMD GPUs.  The data is moved to the GPUs at the start of
the simulation and all computation is done there.  Our GPU offloading
strategy takes advantage of the \amrex\ C++ lambda-capturing
functionality to be performance portable.

We run 3 different types of integration methods: traditional operator
splitting, using Strang splitting, both with integration of the energy
equation and without any temperature evolution during the burn.  \MarginPar{comment about the EOS}.  These methods are compared to the simplified-SDC integtation.

We also vary the CFL number and the integration tolerances.

Show Ni56 mass, intermediate mass

Show time-sequence + compression plot

Show cost plot

@@

Our results suggest that exploring the time-integration method and
integrator tolerances is an important part of demonstrating
convergence of simulations involving explosive reacting flows in
astrophysics.

In the near future we will show how to include various nuclear
statistical equilibrium approximations into the simplified-SDC
formalism while retaining second-order accuracy.

We encourage authors to include a discussion about the reaction integration algorithm
and tolerances in papers to permit comparisons in the future.
